---
title: "R for Reproducible Research: Week 1"
date: "Nov 3, 2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## Instructor Roles ##
# Section                 Instructor  Host    Chat
#----------------------------------------------------
# Workshop Introduction   Randy       Amy     George
# Intro to R and RStudio  Randy       Amy     George
# PM w/ RStudio           George      Randy   Amy
# Seeking Help            Amy         George  Randy
```

## Agenda

Instruction will be based on the [Data Carpentry](https://datacarpentry.org/) workshop, [R for Reproducible Research](http://swcarpentry.github.io/r-novice-gapminder/). This week we will be covering modules 1 - 3:

-   11:00 - Workshop Introduction (Randy)
-   11:10 - [Introduction to R and RStudio](http://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/index.html) (Randy)
-   11:50 - Break
-   12:10 - [Project Management with RStudio](http://swcarpentry.github.io/r-novice-gapminder/02-project-intro/index.html) (George)
-   12:40 - [Seeking Help](http://swcarpentry.github.io/r-novice-gapminder/03-seeking-help/index.html) (Amy)

## Introduction to R and RStudio

### Challenge 1

Which of the following are valid R variable names? Sort them into the corresponding list.

`min_height` `max.height` `_age` `.mass` `MaxLength` `min-length` `2widths` `celsius2kelvin`

-   Valid R variable names

    -   

-   Invalid R variable names

    -   

### Challenge 2

What will be the value of each variable after each statement in the following program?

    mass <- 47.5
    age <- 122
    mass <- mass * 2.3
    age <- age - 20

### Challenge 3

Run the code from the previous challenge, and write a command to compare mass to age. Is mass larger than age?

```{r}
mass <- 47.5
age <- 122
mass <- mass * 2.3
age <- age - 20

# compare mass with age
```

## Project Management with RStudio

### Challenge 1

We’re going to create a new project in RStudio:

1. Click the "File" menu button, then "New Project".
2. Click "New Directory".
3. Click "New Project".
4. Type in the name of the directory to store your project, e.g. "my_project".
5. If available, select the checkbox for "Create a git repository."
6. Click the "Create Project" button.

### Challenge 2

1. Exit RStudio.
2. Navigate to the directory where you created a project in Challenge 1.
3. Double click on the `.Rproj` file in that directory.

### Challenge 3

Download the gapminder data from [here](https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder_data.csv).

1. Download the file (CTRL + S, right mouse click -> "Save as", or File -> "Save page as")
2. Make sure it’s saved under the name gapminder_data.csv
3. Save the file in the data/ folder within your project.

We will load and inspect these data later.

### Challenge 4

It is useful to get some general idea about the dataset, directly from the command line, before loading it into R. Understanding the dataset better will come in handy when making decisions on how to load it in R. Use the command-line shell to answer the following questions:

* What is the size of the file?
* How many rows of data does it contain?
* What kinds of values are stored in this file?

## Seeking Help

### Challenge 1

Look at the help for the `c` function. What kind of vector do you expect you will create if you evaluate the following:

```{r, eval=FALSE}
c(1, 2, 3)
c('d', 'e', 'f')
c(1, 2, 'f')
```

### Challenge 2

Look at the help for the `paste` function. You’ll need to use this later. What is the difference between the `sep` and `collapse` arguments?

### Challenge 3

Use help to find a function (and its associated parameters) that you could use to load data from a tabular file in which columns are delimited with "\t" (tab) and the decimal point is a "." (period). This check for decimal separator is important, especially if you are working with international colleagues, because different countries have different conventions for the decimal point (i.e. comma vs period). hint: use `??"read table"` to look up functions related to reading in tabular data.