---
title: 'Software Carpentry: R for Reproducible Scientific Analysis, Vectorization'
author: "Amy Stonelake"
date: "12/14/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Vectorization

__Questions__

1. How can I operate on all the elements of a vector at once?

__Objectives__

1. To understand vectorized operations in R. 

R functions can operate on all elements of a vector without having to create a loop and iterate over each element.

In this example, the multiplication operation occurs to each element of the vector. 
```
x <- 1:4
x * 2

[1] 2 4 6 8

```

Here's another example where we can add two vectors together. Each element of x is added to the corresponding element of y. 

```
y <- 6:9
x + y

[1] 7 9 11 13

```

compare to doing it with a loop

```
output_vector <- c()
for (i in 1:4) {
  output_vector[i] <- x[i] + y[i]
}
output_vector


[1] 7 9 11 13
```

compare to this:

```
sum_xy <- x + y
sum_xy

[1] 7 9 11 13
```
__Challenge 1__

__Make a new column in the gapminder data frame__ that contains population in units of millions of people. Check the head or tail of the data frame to make sure it worked. 
First let's look at the population data in the gapminder data set.
```
library("gapminder")
head(gapminder$pop)

[1]  8425333  9240934 10267083 11537966 13079460 14880372
```
Now we'll use vectorization to compute the population in units of millions of people. 
```

gapminder$pop_millions <- gapminder$pop/1000000
head(gapminder)

```
Did it work?

__Challenge 2__

On a single graph, plot population, in millions, against year, for all countries. Don't worry about identifying which country is which. 

```{r}
library("gapminder")
library("ggplot2")
gapminder$pop_millions <- gapminder$pop/1000000
ggplot(data = gapminder, mapping = aes(x = year, y = pop_millions)) + geom_point()

```

__Challenge 2, Part 2__

But first, a reminder about the %in% operator...

One of the things you can do with the %in% operator:

1. Use the %in% operator to compare two vectors containing letters or factors (words). 

```
all_of_the_other_reindeer <- c("Dasher", "Dancer", "Prancer", "Vixen", "Comet", "Cupid", "Donner", "Blitzen")

Rudolph <- "Rudolph"

Rudolph %in% all_of_the_other_reindeer

```

What do you get? (FALSE)

Now try this.

```
all_of_the_other_reindeer_plus_Rudolph <- c("Rudolph", "Dasher", "Dancer", "Prancer", "Vixen", "Comet", "Cupid", "Donner", "Blitzen")

Rudolph %in% all_of_the_other_reinder_plus_Rudolph

```
What do you get now? (TRUE)




Now - repeat the graphing exercise above by graphing only for China, India and Indonesia. 

```{r}
library("dplyr")
filter(gapminder, country %in% c("China", "India", "Indonesia")) %>%
ggplot(aes(x = year, y = pop_millions)) + geom_point()

```

__Comparison operators__

Remember...
```
x <- 1:4
x > 2

[1] FALSE FALSE TRUE TRUE

a <- x > 3
a

[1] FALSE FALSE FALSE TRUE

```
Also works with other functions.

```
x <- 1:4
log(x)

[1] 0.0000000 0.6931472 1.0986123 1.3862944
```
*
__Matrix multiplication__

Matrix is created with the matrix() function. Dimensions are defined by passing values for nrow and ncol. 

```
m <-matrix(1:12, nrow=3, ncol=4)
m * -1

Output
     [,1] [,2] [,3] [,4]
[1,]   -1   -4   -7  -10
[2,]   -2   -5   -8  -11
[3,]   -3   -6   -9  -12

```

__Very important__
The * operator is for element-wise multiplication. Matrix multiplication needs the %*% operator. 

```
m %*% matrix(1, nrow=4, ncol = 1)

   [,1]
[1,]   22
[2,]   26
[3,]   30
```

```
matrix(1:4, nrow=1) %*% matrix(1:4, ncol=1)

Output
   [,1]
[1,]   30
```

__Challenge 3__

Given the following matrix.

```
m <- matrix(1:12, nrow=3, ncol=4)
m

Output
    [,1] [,2] [,3] [,4]
[1,]    1    4    7   10
[2,]    2    5    8   11
[3,]    3    6    9   12
```

What is output expected for:

1. m ^ -1
1. m * c(1,0,-1)
1. m > c(0,20)


```
Output
          [,1]      [,2]      [,3]       [,4]
[1,] 1.0000000 0.2500000 0.1428571 0.10000000
[2,] 0.5000000 0.2000000 0.1250000 0.09090909
[3,] 0.3333333 0.1666667 0.1111111 0.08333333

Output
     [,1] [,2] [,3] [,4]
[1,]    1    4    7   10
[2,]    0    0    0    0
[3,]   -3   -6   -9  -12

Output
      [,1]  [,2]  [,3]  [,4]
[1,]  TRUE FALSE  TRUE FALSE
[2,] FALSE  TRUE FALSE  TRUE
[3,]  TRUE FALSE  TRUE FALSE

