---
title: "Introduction to R and RStudio"
author: "Randy Johnson"
date: "11/3/2020"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

Follow along at [Introduction to R and RStudio](http://swcarpentry.github.io/r-novice-gapminder/01-rstudio-intro/index.html).

* Questions
    * How to find your way around RStudio?
    * How to interact with R?
    * How to manage your environment?
    * How to install packages?

* Objectives
    * Describe the purpose and use of each pane in the RStudio IDE
    * Locate buttons and options in the RStudio IDE
    * Define a variable
    * Assign data to a variable
    * Manage a workspace in an interactive R session
    * Use mathematical and comparison operators
    * Call functions
    * Manage packages

## Arithmetic

```{r arithmetic}
# calculator
1 + 100 # comment

3 + 5 * 2

# order of operations is important
(3 + 5) * 2

3 + (5 * 2)

# scientific notation
4 / 100000
4 / 1e5

```

## Mathematical functions

```{r math functions}

sin(1)

log(10)
log10(10)

exp(.5)

```

## Comparing things

```{r comparing things}

# is 1 equal to 1?
1 == 1

# is 1 not equal to 2?
1 != 2
3 == 2

# is 1 less than 2?
1 < 2

# is 1 less than or equal to 2?
1 <= 2

# is 1 greater than -9?
1 >= -9


# is 1.1 equal to 1.1 + 1e-15
all.equal(1.1, 1.1 + 1e-15)
```

## Variables and assignment

```{r variables}
# create a variable
x <- 1/40

log(x)
log10(x)

# change x
x <- 100

# increment x by 1
x <- x + 1

# create a new variable with x
y <- x*3

# change x again
x <- x + 1

celsius2kelvin <- 1
```

### Challenge 1 solution (Poll a)

Which of the following are valid R variable names?

* `min_height`
* `max.height`
* `_age`
* `.mass`
* `MaxLength`
* `min-length`
* `2widths`
* `celsius2kelvin`

## Vectorization

```{r vectorization}
# create a vector from 1 to 5
x <- 1:5

# square x
x^2

# log of x
log(x)
```

## Managing your environment

```{r environment}

# what is in my environment
ls()

# remove min_age
rm(y)

# is min_age still there?
ls()

# remove everything
rm(list = ls())
```

## R Packages

```{r packages}
# load dplyr
library(dplyr)

# run the base version of intersect, not dplyr version
# base::intersect()

# install packages via tools>install packages or...
#install.packages('dplyr')
```

## Challenge Problems

### Challenge 2 solution

What will be the value of each variable after each statement in the following code?

```{r challenge2}
# we wrap each statement in () to print the result *and* save it to a variable
(mass <- 47.5)
(age <- 122)
(mass <- mass * 2.3)
(age <- age - 20)
```

### Callenge 3 solution (Poll b)

Following up from the previous challenge, write a command to compare mass to age. Is mass larger than age?

* Yes
* No

```{r challenge3}

```

### Challenge 4 solution

Clean up your working environment by deleting the mass and age variables.

```{r challenge4}

```

### Challenge 5 solution

Install the following packages: `ggplot2`, `dplyr`, `gapminder`

```{r challenge5, eval=FALSE}
# We could do this in RStudio with Tools > Install Packages...
# or we could install them from the command line as follows:

```
